---
title: "IDA_Assignment2"
author: "Yuxin Yang"
date: "2023-03-21"
output: pdf_document
---

## Question 2(b)

```{r}
require(maxLik)
#load("~/Desktop/dataex2.Rdata")
load("dataex2.Rdata")
#Log likelihood function set to maximized
log_like = function(param, data) {
  x = data[,1]; r = data[,2]
  mu = param
  #dnorm density function, pnorm cumulative distribution function
  sum(r*dnorm(x, mean=mu, sd=1.5, log=TRUE) + 
               (1 - r)*pnorm(x, mean=mu, sd=1.5, log.p=TRUE))
}
#Get MLE
mle = maxLik(logLik = log_like, data = dataex2, start = c(mu = 1))
summary(mle)
```

The Newton-Raphson maximisation method is used to optimize the likelihood, $\widehat{\mu}_{\text{MLE}} =  5.53$.

We will now do the same but using the optim function.
```{r}
res <- optim(c(1), log_like, data = dataex2, control = list(fnscale = -1), 
             hessian = TRUE)
res
```

## Question 4
```{r}
require(maxLik)
#load("~/Desktop/dataex4.Rdata")
load("dataex4.Rdata")
#Replace NA in Y into 2
dataex4 <- replace(dataex4, is.na(dataex4), "2")
dataex4$Y <- as.numeric(dataex4$Y)
#Create a column Z to indicate if it is missing value in Y
Z <- dataex4$Y
#If it is missing value, Z <- 0, else, Z <- 1
Z[Z == 0] <- 1
Z[Z == 2] <- 0
dataex4$Z <- Z

#Probability function
prob <- function(beta, x) {
  return(exp(beta[1] + x*beta[2])/(1 + exp(beta[1] + x * beta[2])))
}
#E-step
#Q function
Q_function <- function(param, data){
  beta0 = param[1]; beta1 = param[2]
  sum(data$Y * data$Z * (beta0 + beta1*data$X) - log(1+exp(beta0 + beta1*data$X)) 
      + (1-data$Z) * (beta0 + beta1*data$X) * prob(ex_beta, data$X))
}
#For stop criterion, we use tol <- 1e-5
tol <- 1e-5
ex_beta <- c(0,0)
repeat{
  beta <- coef(maxLik(Q_function, data = dataex4, start = ex_beta))
  if (max(abs(beta-ex_beta)) < tol){
    break
  }
  ex_beta <- beta
}
beta
```

The results are $\widehat{\beta_0}_{\text{MLE}} = 0.976$ and 
$\widehat{\beta_1}_{\text{MLE}} = -2.48$ for 
$\boldsymbol{\beta} = (\beta_0, \beta_1)'$.

## Question 5(b)
```{r}
#load("~/Desktop/dataex5.Rdata")
load("dataex5.Rdata")
em.mixture.two.dis <- function(y, theta0, eps){
  n <- length(y)
  theta <- theta0
  p <- theta[1]
  lambda <- theta[2]
  mu <- theta[3]
  diff <- 1
  while (diff > eps) {
    theta.old <- theta
    
    #E-step
    ptilde1 <- p * (lambda * y^(-lambda-1))
    ptilde2 <- (1 - p) * (mu * y^(-mu-1))
    ptilde <- ptilde1/(ptilde1 + ptilde2)
    
    #M-step
    p <- mean(ptilde)
    lambda <- sum(ptilde)/sum(ptilde * log(y))
    mu <- sum(1 - ptilde)/sum((1 - ptilde) * log(y))
    
    theta <- c(p, lambda, mu)
    diff <- sum(abs(theta - theta.old))
  }
  return(theta)
}

res <- em.mixture.two.dis(y = dataex5, c(0.3, 0.3, 0.4), 0.0001)
p <- res[1]
lambda <- res[2]
mu <- res[3]

p; lambda; mu
```

The MLEs for each component of $\boldsymbol{\theta}$ are:
$\widehat{p}_{\text{MLE}} = 0.79$;
$\widehat{\lambda}_{\text{MLE}} = 0.97$;
$\widehat{\mu}_{\text{MLE}} = 6.67$.

The histogram of the data with the estimated density superimposed is as follows:
```{r}
hist(dataex5, breaks = "FD", main="Random Sample from Mixture Distribution",
     xlab = "Y",
     ylab = "Density",
     cex.main = 1.5,
     cex.lab = 1.5, 
     cex.axis = 1.4,
     xlim= c(0,20),
     ylim= c(0,1),
     freq = FALSE)
# Superimposing mixed density distribution
curve(p * (lambda * x^(-lambda-1)) + (1 - p) * (mu * x^(-mu-1)), add= TRUE, 
      col= "blue2",lwd=1.5)
```
